A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2014-02-05 response/response.tex

\documentclass{article}
\usepackage[left=1.0in,right=1.0in,top=1.0in,bottom=1.0in]{geometry}
\usepackage{fancyhdr}
\usepackage{color}
\usepackage{amsmath}
\definecolor{light-gray}{gray}{0.5}

\begin{document}

\begin{center}
    {\LARGE \bf Point-by-Point Response to Reviewer's Comments} \\
     Author's Responses in {\color{red} Red}
\end{center}

%%%% First Reviewer %%%%
\section{Reviewer}
%
\subsection*{General Statement}
In general, I am not convinced of the benefits of particle methods when compared to traditional finite element analysis. Although they eliminate the sometimes troublesome problem of mesh generation, they create new problems of their own. In particular, they are hard to debug (because the patch test often doesn’t work), it is typically difficult to pin down an order of accuracy, and it is typically difficult to apply boundary conditions. Nonetheless, I recognize that they are part of the toolbox and therefore are within the scope of IJNME.
%
{\color{red}
  \begin{itemize}
    \item We have added a paragraph at the end of Section 1 that highlights the difference between peridynamics and previous “particle methods.” The main difference is that in peridynamics, we numerically approximate a different set of continuum equations.
  \end{itemize}}

This particular paper, however, has insufficient novelty to merit publication in its present form. In particular,

\begin{enumerate}
  \item The peridynamic framework, including the method to compute the deformation gradient, is already established in the previous literature.
  {\color{red}
    \begin{itemize}
      \item We have added statements in the abstract and introduction that highlight what is new about the work being reported. Specifically, this represents the first successful implementation of a rate-dependent material model within peridynamics. We also present the first full-scale 3D solution of a realistic, reasonably complex application. This work therefore demonstrates to the community, for the first time, the ability of the peridynamic method to solve practical problems involving complex materials resulting in good agreement with experimental data.
    \end{itemize}}
  %
  \item The viscoplastic formulation and constitive model follows fairly classical nonlinear solid mechanics, and therefore does not appear to be novel. 
  {\color{red}
    \begin{itemize}
      \item Viscoplasticity has never been used within peridynamics.  See previous responses.
    \end{itemize}}
  %
  \item The computational studies are quite limited, by the authors’ own admission (“while failure can be modeled with peridynamics, it is beyond the scope of this paper; therefore, these experiments were not attempted in numerical simulation”, p. 19). 
  {\color{red}
  \begin{itemize}
    \item The focus of this paper was to show that peridynamics could be used with conventional (stress-strain) constitutive models.   Modeling failure is the main advantage of the peridynamic approach, which has been demonstrated in the cited papers and will be further developed in a future paper with the combination of a viscoplastic constitutive model and a novel failure mechanism applied by breaking ``bonds''.
  \end{itemize}}
  %
\end{enumerate}

\subsection*{Specific Comments}

\begin{enumerate}
%
   \item p. 3, It has been proved that Equation 1 reduces to the classical continuum partial differential equation\dots Which one? Momentum?
   {\color{red}
     \begin{itemize}
        \item Yes, linear momentum.  This has been made more clear in the rewritten introduction.
     \end{itemize}}
%
  \item p. 5, should be $F = I + \nabla_x u$ (standard notation)
   {\color{red}
     \begin{itemize}
        \item The notation I used is identical to the notation used by Lawrence E. Malvern in his classic book \underline{Introduction to the Mechanics of a Continuous Medium} (p.\ 162 in the version I have).  This is further clarified by showing the incidental notation in the revision.
     \end{itemize}}
%
  \item Eq. (30) is an additive decomposition of the deviator of the referential tensor $\mathbf{d}$ based on the additive decomposition of the rate of deformation $\mathbf{D}$. A reference and/or some background should be included.
  {\color{red}
     \begin{itemize}
        \item Eq.\ (30) is an additive decomposition of the deviatoric part of $\mathbf{d}$ which is referred to as the {\em unrotated rate of deformation tensor}.  $\mathbf{d}$ is found from Eq.\ (17) by transforming $\mathbf{D}$, the {\em rate of deformation tensor}, with the rotation tensor $\mathbf{R}$.  This entire process is taken from the paper by Flanagan and Taylor ``An accurate numerical algorithm for stress integration with finite rotations'', which I believe to be a standard algorithm for developing kinematics in large displacement problems.
     \end{itemize}}
%
     \item Notation: Usually lower case letters are used for spatial quantities and upper case letters are used for referential quantities, e.g., $\mathbf{d}$ is standard for the rate of deformation tensor.
  {\color{red}
     \begin{itemize}
        \item All of the notation used in this paper follows very closely the notation used by Malvern and/or Flanagan and Taylor.
     \end{itemize}}
%
     \item I don't understand why eq.\ (35) is presented. As far as I know, the use of previous values of strain in order to extrapolate the strain rate is never used in the literature, and all previous authors obtain the strain by solving an equation like (36). So unless there is some good reason to believe that (35) might be plausible (e.g., an instance where it is successfully used in the previous literature), I would drop it from the paper.
  {\color{red}
     \begin{itemize}
        \item For the reviewer's reference a paper from Brown University by Peirce, Shih, and Needleman titled ``A tangent modulus method for rate dependent solids'' uses a method similar to what is presented in Eq.\ (35), a fully explicit method.
        \end{itemize}}
%
        \item Mistaken forward reference to eq. 59 on p. 14.
  {\color{red}
     \begin{itemize}
        \item Corrected in revision
     \end{itemize}}
%
      \item The paper mixes up modeling with discretization. For example, the model apparently doesn't discuss how the pressure component of stress is handled, and yet this pops up in the discussion of the numerics (eq.\ (51)).
  {\color{red}
     \begin{itemize}
        \item I do not fully understand this comment, von Mises plasticity only deals with the deviatoric part of stress to define a yield surface, but the pressure component must be added back in after the deviatoric stress has been updated. Nevertheless, a mistake was found in Eq.\ (51) and corrected in the revision.
      \end{itemize}}
%
        \item The total strain is apparently defined on a particle-by-particle basis using (61) and (12). I suppose the plastic strain is a nodal history variable?
  {\color{red}
     \begin{itemize}
        \item All stress or strain quantities including the internal state variable, {\em equivalent plastic strain}, are calculated for each particle at each timestep. 
     \end{itemize}}
%
        \item The number of nodes required seems to be huge in both simulations in Section 10. Is there any compensating benefit (versus finite elements)?
  {\color{red}
     \begin{itemize}
        \item The number on nodes was large, but these simulations ran in only a few hours on 4 processors.  The largest compensating benefit is the ability to model failure (outside the scope of this paper, but discussed in others) with the same set of fundamental equations without the need for separate governing equations for crack initiation, propagation, etc. (e.g., cohesive zones, element death).
     \end{itemize}}
 %
        \item The literature review includes previous papers (mainly by the same group of authors) on peridynamics but does not review the larger universe of particle methods for solid mechanics, and I think some references in this regard would be helpful. In particular, I think some of the previous particle methods are quite similar to peridynamics. Unfortunately, I am not an expert in this, but I know of at least three previous papers proposing particle methods for solid mechanics.
       \begin{enumerate}
         \item Rabczuk, Belytschko and Xiao, Stable particle methods based on Lagrangian kernels, CMAME 193 (2004) 1035-1063.
         \item Delaplace and Ibrahimbegovic, Performance of time-stepping schemes for discrete models in fracture dynamics, IJNME 65 (2005) 1527-1544.
         \item Gao and Klein, Numerical simulation of crack growht in an isotropic solid with randomized internal cohesive bonds, JMPS 46 (1998) 187-218.
       \end{enumerate}
%
     {\color{red}
     \begin{itemize}
        \item The main reason other particle methods were not reviewed was because all of these other methods are based on theory which solves partial differential equations and are therefore fundamentally different from the nonlocal integral equations solved in peridynamics.  However, additional discussion was added in the revision which points out these differences and references the reviewers suggested papers where appropriate. 
         \end{itemize}}
        %
        \item Inconsistent verb tense in second paragraph of summary (is, was.)

p. 2, Where $\rho \rightarrow$ where $\rho$

Van der Walls $\rightarrow$ Van der Waals

any pair of particles interact $\rightarrow$ interacts

singe $\rightarrow$ single

spacial $\rightarrow$ spatial

principle $\rightarrow$ principal

et. al $\rightarrow$ et al. (italicize)

%
     {\color{red}
     \begin{itemize}
       \item Corrected in revision
     \end{itemize}}
%
\end{enumerate}

\section{Reviewer}
\subsection*{General Statement}
I recommend that the authors prepare a revised version of their paper to be considered for publication in the International Journal of Numerical Methods in Engineering.  Specifically, the authors should consider adding discussion related to the following eleven comments (1)-(11).
%
\subsection*{Specific Comments}

\begin{enumerate}

  \item This is the first time that I have read about the peridynamic theory.  In many ways it seems to be similar to the notions used in SPH (smooth particle hydrodynamics).  It is know that the SPH theories have difficulties handling material strength in a consistent stable manner.  I would expect the same problems with the peridynamic theory. It would be helpful to compare these theories and explain important differences and potential advantages. 
%
     {\color{red}
     \begin{itemize}
        \item Peridynamics is fundamentally different from SPH because SPH is still based on theory that solves partial differential equations.  In SPH, properties of each particle are smeared over a smoothing length, but the motion of the particles is still governed by PDE's.  In peridynamics the conservation of momentum equation is a nonlocal integral equation capable of handling even discontinuous displacements fields that would cause problems in SPH where the PDE's could not be evaluated.  SPH also suffers from a tensile instability that is not problematic for peridynamics.  Additional discussion has been added to the introduction to address these issues.
     \end{itemize}}
%
  \item Introducing nonlocality through the integral over the sphere of radius $\delta$ in the reference configuration violates the notion of a finite wave speed in the material.  Obviously the same problem occurs when finite elements are used to average deformation information over a finite region.  Some comments about the effect of this in this peridynamic theory would be helpful
     {\color{red}
     \begin{itemize}
        \item As the reviewer points out this is a similar issue in finite elements.  Wave speeds have been analyzed in peridynamics and have been shown to not have an infinite wave speed, but the wave speeds are a function of wave length.  When analyzed it turns out that the bulk modulus is directly related to $\delta$ and therefore, becomes an integral part of the defination of wave speed, i.e. $\delta$ is a material property.
     \end{itemize}}
%
  \item On page 2 it is stated that:
  \begin{quotation}
  \noindent ``In this implementation, $\omega$ is always either 0 or 1, which corresponds to whether the bond is `broken' or not.'' 
 \end{quotation}
 It seems to me that $\delta$ defines the radius of influence of material points and $\omega$ defines the strength of the influence.  In particular, it is surprising to me that a uniform distribution of $\omega$ with a sharp cut-off would not cause serious problems with wave propagation.  It would be helpful to comment on this in a revised version of the paper. 
     {\color{red}
     \begin{itemize}
        \item The reviewer is correct in his recognition of $\omega$ as a strength of influence.  So far only uniform distributions of $\omega$ have been evaluated and they do not show any numerical instablity.  Additional discussion was added to the revision.
     \end{itemize}}
%
  \item It would be helpful to provide a discussion of how displacement and traction boundary conditions can be implemented using this theory.
     {\color{red}
     \begin{itemize}
        \item Displacements, velocities, and body force densities can be applied to regions of finite volume.  Information on this was added to the revised article in the discussion of the numerical simulations.  See Silling ``Reformulation of elasticity theory for discontinuities and long-range forces'' cited in article for further reading.
     \end{itemize}}
%
  \item It would be helpful to discuss the specific approximations that are used when the sphere of influence (associated with $\delta$) extends outside of the boundaries of the body. 
     {\color{red}
     \begin{itemize}
        \item There are no issues when $\delta$ extends outside the boundaries of the body.  A proof has been included which shows that the deformation gradient can be correctly calculated for any assumed deformation state in the revised article.
     \end{itemize}}
 %
  \item It is my understanding that this method tracts the motion of material points. Since the sphere of influence (associated with $\delta$) is defined relative to the reference configuration, the same material points remain in communication even if the material is highly stretched or highly compressed. It would seem to me that for the highly stretched case the method would yield unphysical dispersion of waves since the region of influence can be quite large.
     {\color{red}
     \begin{itemize}
        \item The reviewer's assumptions are correct, but the same thing could happen in a finite element simulation as well when elements become severely stretched or distorted.  It would be possible to redefine the $\delta$ at every time step where material points far away would be removed from this horizon and others added; however, this adaptivity feature does not currenlty exist in the code used here, Emu.
     \end{itemize}}
 %
  \item For an elastic solid, which has a unique unstressed reference shape, it makes sense to refer the constitutive response to a reference configuration.  In contrast, the material response of an elastic-viscoplastic material, which has no unique unstressed configuration or shape, should be formulated in terms of constitutive equations which are explicitly independent of the arbitrary choice of the reference configuration. Moreover, since the main motivation for developing meshless methods is to handle large deformations it is unfortunate that the elastic-viscoplastic constitutive equations have been formulated with respect to the reference configuration using the rotation tensor $\mathbf{R}$ in (15).  Also, the integral operators are specified by considering regions defined in the reference configuration.  It is possible to develop this method in an Eulerian framework more similar to SPH methods? 
     {\color{red}
     \begin{itemize}
       \item The method of using $\mathbf{R}$ to provide an unrotated configuration in which to apply the constitutive model such that the stress rates will be objective is used in many finite element codes, including the production code at Sandia National Laboratories, PRESTO.  This method is described in detail in the Flanagan and Taylor paper ``An accurate numerical algorithm for stress integration with finite rotations.''  Futhermore, Eulerian frame constitutive equations can be used in peridynamics as long as we define $\mathbf{F}$ and $\mathbf{R}$.
     \end{itemize}}
     %
  \item The constitutive equations used for determining viscoplastic response are not clear. It is my understanding that (28) is a yield function which limits the magnitude of the deviatoric stress during loading
  \begin{equation}
    S = | \mathbf{S} | = \sqrt{\frac{2}{3}} G \quad \mbox{during loading.}
    \tag{28}
  \end{equation}
  Then, using the additive decomposition (29), the elastic constitutive equation (31), the flow rule (32) and (33), and the expression (38) it can be shown that 
  \begin{equation}
    \dot{S} = 2 \mu \left ( \mathbf{Q} : \dot{\mathbf{e}} - \sqrt{\frac{3}{2}} \dot{\varepsilon}_p \right ) = \sqrt{\frac{2}{3}} \dot{G}
    \tag{A1}
  \end{equation}
  However, viscoplasticity is introduced in by taking the value of $G$ to be a function of the equivalent plastic strain $\varepsilon_p$ and plastic strain rate $\dot{\varepsilon}_p$
  \begin{equation}
    G = f \left ( \varepsilon_p ,  \dot{\varepsilon}_p \right ) .
    \tag{34}
  \end{equation}
  Thus, substituting (34) into (A1) yields
  \begin{equation}
    2 \mu \left ( \mathbf{Q} : \dot{\mathbf{e}} - \sqrt{\frac{3}{2}} \dot{\varepsilon_p} \right ) = \left ( \frac{\partial f}{\partial \varepsilon_p} \right ) \dot{\varepsilon}_p + \left ( \frac{\partial f}{\partial \dot{\varepsilon}_p} \right ) \ddot{\varepsilon}_p
    \tag{A2}
  \end{equation}
  which is a second order differential equation for the equivalent plastic strain.  It is not clear that the proposed procedure described in sections 6 and 7 actually solve this second order equation for $\varepsilon_p$.
     {\color{red}
     \begin{itemize}
       \item What is derived in the paper is the finite differencing scheme for solving for an increment of equivalent plastic strain.  There is no need to consider the continous derivatives when we do this.  Nevertheless I will start with equation (A1) and show that it equals the final result from the paper.  Taking the right-hand equality from (A1) we can substitute (39)\footnote{\color{red} Equation numbers refer to equation numbers in original paper, not the revision} for $\mathbf{Q:\dot{e}}$. Noting that because $\mathbf{Q}$ and $\mathbf{\dot{e}}$ are symmetric tensors $\mathbf{Q:\dot{e}} = \mathbf{\dot{e}:Q}$.
        \begin{equation}
	  2 \mu \left ( \dot{e} - \sqrt{\frac{3}{2}} \dot{\varepsilon}_p \right ) = \sqrt{\frac{2}{3}} \dot{G}
	    \tag{B1}
	  \end{equation}
	  Now substitute (38) and (41)
	  \begin{equation}
	    2 \mu \dot{e}^e = \sqrt{\frac{2}{3}} \dot{G}
	    \tag{B2}
	  \end{equation}
	  Convert from continous derivatives to rates.
	  \begin{equation}
	    2 \mu \frac{\Delta e_{t-\Delta t/2}^e}{\Delta t} = \sqrt{\frac{2}{3}} \frac{\Delta G_{t-\Delta t/2}}{\Delta t}
	    \tag{B3}
	  \end{equation}
	  Multiply by $\Delta t$ and expand the RHS of (B3)
	  \begin{equation}
	    2 \mu {\Delta e_{t-\Delta t/2}^e} = \sqrt{\frac{2}{3}} \left ( G_{t}- G_{t-\Delta t} \right )
	    \tag{B4}
	  \end{equation}
	  and of course $G_t = \sqrt{3/2} S_t$ and similary for $G_{t-\Delta t}$, using these in (B4)
	  \begin{equation}
	    2 \mu {\Delta e_{t-\Delta t/2}^e} = S_t - S_{t-\Delta t}
	    \tag{B5}
	  \end{equation}
	  finally substituting the rate form of (41) and rearranging we have:
	  \begin{equation}
	    \Delta e_{t-\Delta t/2} - \frac{S_t - S_{t-\Delta t}}{2 \mu} - \Delta \lambda_{t-\Delta t/2} = 0
	    \tag{B6}
	  \end{equation}
	  (B6) is exactly (44).  The rest of derivation follows the paper from (44).
     \end{itemize}}
 %
  \item Line 7 after (8).  $\mathbf{K}$  is a second order moment not a first order moment since $\boldsymbol{\xi} \otimes \boldsymbol{\xi}$ is a quadratic term.
     {\color{red}
     \begin{itemize}
        \item The description of $\mathbf{K}$ as a moment has been removed in the revision.
     \end{itemize}}
     %
  \item The value of $\delta$ used for the simulations should be specified.  In this regard, it would be good to also state how many material points used in the simulation fit into this sphere of radius $\delta$. 
     {\color{red}
     \begin{itemize}

### 2014-02-05 response/response.tex

        \item Suggestion was included in revision along with additional discussion about how changing $\delta$ makes insignificant changes in the solution of the Taylor impact problem.
     \end{itemize}}
        %
  \item Figure 2 refers to true strain but the measure $\mathbf{E}$ in (61) is the Lagranian strain.  Please provide a definition of the strain measure being used for true strain in Fig. 2. 
     {\color{red}
     \begin{itemize}
        \item Suggestion was included in revision.
     \end{itemize}}
\end{enumerate}

\end{document}

### 2014-02-05 jacobian_calc_CMAME.tex
Added response folder and trackchanges.sty

\begin{document}

@@

The methods were compared by running two sets of test problems where each method would solve a problem concurrently. Each test problem simulated a $\SI{4}{\meter}$ by $\SI{0.5}{\meter}$ by $\SI{0.5}{\meter}$ meter block of a material undergoing tension along the axis parallel to the long dimension of the block. The test problems were set up in \emph{Perdigm} as quasistatic equilibrium problems, where the displacement boundary conditions used to apply tension were applied gradually, in `load steps', and an equilibrium solution for each load step was achieved before applying the next load step. The purpose of applying the load in gradual steps is a widely used technique to ensure the Newton's method is always initialized near the actual solution and therefore likely to converge. The fictional material represented in the model had values of $\SI{1.515e4}{\mega\pascal}$ for bulk modulus, and $\SI{7.813e4}{\mega\pascal}$ for shear modulus. The peridynamic horizon, a length scale that effectively sets the bandwidth of tangent-stiffness in the computation, used in the test problems was always set to three times the nominal discretization size (i.e. node spacing in the particle discretization scheme). This results in a minimum tangent-stiffness matrix bandwidth of 7 (for an optimal node numbering scheme).

A series of nine single compute core (i.e. serial) test problems were run with the discretization level being refined with every test in the series. The aim of increasing the refinement was to see if possible differences between the methods in terms of accuracy and speed would arise. This series would comprise the single core runs. The specific parameters of these tests can be found in \emph{Peridigm} input {\tt xml} files included in the archive for this paper which can be found in the corresponding author's website. These {\tt xml} files allow users with appropriately modified versions of \emph{Peridigm} to reproduce the results found in this paper.

Another series of four test problems were simulated, however this time the number of compute cores used to solve the problem was increased from test to test while the discretization level was held constant at 1 million computational nodes (3 million degrees of freedom). This allowed for sufficient computational nodes per core even at the highest level of parallelization such that message passing computation did not overwhelm the simulation.   The specific parameters of these tests can be found in \emph{Perdigm} input {\tt xml} files also included in the archive for this paper on the corresponding author's website.

@@

It would be a poor comparative study to compare tangent-stiffness matrices from different problems, load steps that start with different current configurations, or from different iterations; because of this, it was necessary to solve one
load step and conclude each Newton iteration within that load step by updating the displacement iterate with only the results of the AD method and to subsequently feed all four methods the same updated displacement in the following iteration. This decision precluded a comparison of Newton iteration convergence rate, since if the methods were allowed to solve a problem at their individual pace, differences in accuracy would produce differences in guess updates and therefore the number and nature of Newton iterations performed (e.g. lower accuracy predictions of the algorithmically correct tangent-stiffness could cause a loss of quadratic convergence).  Different guesses from iterations started with different previous guesses could not be data for a valid comparison of tangent-stiffness accuracy between methods.  Additionally, having each of the methods physically operate in the same process, serial or one of associated parallel, allowed the comparison of tangent-stiffness matrix calculation time as cached from random access memory rather than from the hard-disk.  Caching from RAM gives the benefit of vastly greater speed and the simpler programming compared to some other solution speculatively involving dynamic file management on files which for one of the components of the study would be on the order of 10 terabytes in size. However, the price of running the methods together in the same process and avoiding the use of the hard-disk was that at least two tangent-stiffness matrices had to be stored in RAM during the simulation, which meant that special high-memory compute nodes were needed for the 1 million peridynamic node test series.

The other goal was to compare the four methods, that is including AD, on the basis of speed. Speed is defined as the total compute-time of one iteration of a Newton step which includes the tangent-stiffness calculation and parallel assembly. The speed of iteration was measured because it could be so done at the same time as accuracy was being measured given a single tangent-stiffness calculation.  It was assumed that the order that each method was evaluated in parallel was unimportant, that evaluating each method successively within each solver iteration did not affect their performance individually and that speed of computation did not change with time. These assumptions allowed the test program to run the same problem with each of the methods at effectively the same time and generate an equal volume of data from each method. It was also assumed that the tangent-stiffness matrix calculation time for each of the methods did not vary based on the current configuration, as it changes slightly from iteration to iteration as the Newton solution iterate is updated.  This assumption allowed calculation time measurements to be averaged over all iterations within a single load step. The purpose of averaging calculation time measurements was to informally address the extraneous variables of parallel evaluation order and computer system load due to system processes not associated with the study. While not part of the study, the example programs available on the corresponding author's website allow the reader to make a comparison of the methods for themselves on the basis of accuracy and convergence rate for two example nonlinear systems. In these examples, one will notice that the finite difference methods will lose quadratic convergence when the step size selected too large such that accuracy relative to AD is decreased.

Once the quantities-of-interest (QoIs) for comparison were determined, the task was to find a way to take measurements that satisfied our goals. The goals of collecting accuracy and speed measurements were achieved by developing and implementing metrics within the simulation program used in the study. The metric used to measure the accuracy of a tangent-stiffness matrix was the Frobenius norm (analogous to an $l^2$-norm except defined on a matrix) of the element-wise difference between the tangent-stiffness matrix produced by the method being evaluated and the tangent-stiffness matrix produced by the AD based method (i.e., the \emph{exact} derivatives), given that both methods were set upon the same problem. The lower the value of this metric, the closer the tangent-stiffness of the method being evaluated (i.e, CS, AD, FD)  was to the AD Jacobian, the more accurate the method was. The expression for the accuracy metric is
%
\begin{equation} 
  D = \sqrt{\sum_i \sum_j(K^{AD}_{ij} - K^{M}_{ij})^2}
\end{equation}
%
Where $D$ is distance, $K^{AD}$ is the tangent-stiffness matrix produced by the AD based method, $M$ is replaced with either CS, FD, or CD depending on the method being compared.  The metric used to compare the speed of the different methods was the time in seconds required to calculate the tangent-stiffness matrix. A final basis of comparison used in the study called \emph{computational efficiency} is based on the ratio of calculation time to the total number of tangent-stiffness matrix elements.

\subsubsection{Computers and other software} 
%
Discretization for the test problems run in the study was done using the software package \emph{Cubit} developed at Sandia National Laboratories \cite{ref-Cubit}. \emph{Cubit} generates a finite element mesh that is internally converted to a ``particle'' discretization internally when calling \emph{Peridigm}, its important to node that the number of computational nodes in \emph{Peridigm} correspond to the number of finite elements in the discretization, not the number of finite element nodes.  An example journal script can be found on the corresponding author's website.

The test problems were run the \emph{Stampede} HPC cluster computer housed at \emph{TACC} (Texas Advanced Computing Center) at The University of Texas at Austin. The single core test series was run using the {\tt normal} queue compute nodes, while the parallel test series was run using one to four {\tt large memory} nodes each having one terabyte of local RAM. This large amount of memory was required for reasons explained in Section~\ref{JGAM}.

\subsubsection{Data Reduction} 
The output produced by \emph{Perdigm} including the additional accuracy and speed measurements was redirected from the console to text files by the resource manager, \emph{SLURM} (Simple Linux User Resource Manager).  Directories and filenames were chosen so that data would be easy to sort by individual test run.  Extraneous information was manually deleted from copied versions of the output text files so that they would be regular enough to be easily read by scripts written in \emph{Python}. These scripts averaged accuracy or speed data for a tests corresponding to a particular peridynamic node density over all iterations for that run, and then plotted this averaged data  as a function of peridynamic node density as it varied from test to test in the series.  Similarly, for the parallel tests, the data was averaged and plotted in much the same way, but as a function of number of compute cores used to solve the problem rather than as a function peridynamic node density which was held constant. Confidence intervals were not calculated on the measurements. The main reason is that network traffic from other users makes running tests on the HPC cluster a time varying process. These data reduction and plotting scripts can be found on the corresponding author's website.

\section{Results and Discussion}
%
This section contains the average results from the study presented in tabular and graphic format.  Table~\ref{tab:results} shows the compute core count, number of nonzero tangent-stiffness (TS) non-zero matrix elements, load step average calculation time and load step average accuracy. The units for accuracy are derived from the
expression used to calculate the TS matrix elements and the units used to define bulk material properties, mentioned in Section~\ref{tcs}.
%
\note[JTF]{Michael, work on cleaning this table up for legibility}
\note[Michael]{Columns are now separated out and can be fine tuned}

\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

>>>>

\begin{document}\sloppy

@@

The methods were compared by running two sets of test problems where each method would solve a problem concurrently. In this context, concurrently means successively, drawing from the same independent variables, yet within the same computational process; this definition is characterized in further detail in Section~\ref{JGAM}. Each test problem simulated a $\SI{4}{\meter}$ by $\SI{0.5}{\meter}$ by $\SI{0.5}{\meter}$ meter block of a material undergoing tension along the axis parallel to the long dimension of the block. The test problems were set up in \emph{Perdigm} as quasistatic equilibrium problems, where the displacement boundary conditions used to apply tension were applied gradually in ``load steps'', and an equilibrium solution for each load step was achieved before applying the next load step. The purpose of applying the load in gradual steps is a widely used technique to ensure the Newton's method is always initialized near the actual solution and therefore likely to converge. The fictional material represented in the model had values of $\SI{1.515e4}{\mega\pascal}$ for bulk modulus, and $\SI{7.813e4}{\mega\pascal}$ for shear modulus. The peridynamic horizon, a length scale that effectively sets the bandwidth of tangent-stiffness in the computation, used in the test problems was always set to three times the nominal discretization size (i.e. node spacing in the particle discretization scheme). This results in a minimum tangent-stiffness matrix bandwidth of 7 (for an optimal node numbering scheme).

A series of nine single compute core (i.e. serial) test problems were run with the discretization size being refined with every test in the series. The aim of increasing the refinement was to examine differences between the methods in terms of accuracy and speed. This series would comprise the single core runs. The specific parameters of these tests can be found in \emph{Peridigm} input {\tt xml} files included in the paper repository. These {\tt xml} files allow users with the appropriately modified version of \emph{Peridigm} to reproduce the results found in this paper.

Another series of four test problems were simulated, however this time the number of compute cores used to solve the problem was increased from test to test while the discretization level was held constant at 1 million computational nodes (3 million degrees of freedom). This allowed for sufficient computational nodes per core even at the highest level of parallelization such that message passing computation did not overwhelm the simulation.   The specific parameters of these tests can be found in \emph{Perdigm} input {\tt xml} files also included in the paper repository.

@@

It would be a poor comparative study to compare tangent-stiffness matrices from different problems, load steps that start with different current configurations, or from different iterations; because of this, it was necessary to solve one
load step and conclude each Newton iteration within that load step by updating the displacement iterate with only the results of the AD method and to subsequently feed all four methods the same updated displacement in the following iteration. This decision precluded a comparison of Newton iteration convergence rate, since if the methods were allowed to solve a problem at their individual pace, differences in accuracy would produce differences in guess updates and therefore the number and nature of Newton iterations performed (e.g. lower accuracy predictions of the algorithmically correct tangent-stiffness could cause a loss of quadratic convergence).  Different guesses from iterations started with different previous guesses could not be data for a valid comparison of tangent-stiffness accuracy between methods.  Additionally, having each of the methods physically operate in the same process, serial or one of associated parallel, allowed the comparison of tangent-stiffness matrix calculation time as cached from random access memory rather than from the hard-disk.  Caching from RAM gives the benefit of vastly greater speed and simpler programming compared to some other solution speculatively involving dynamic file management on files which for one of the components of the study would be on the order of a terabyte in size. However, the price of running the methods together in the same process and avoiding the use of the hard-disk was that at least two tangent-stiffness matrices had to be stored in RAM during the simulation, which meant that special high-memory compute nodes were needed for the 1 million peridynamic node test series.

The other goal was to compare the four methods, including AD, on the basis of speed. Speed is defined as the total compute-time of one iteration of a Newton step which includes the tangent-stiffness calculation and parallel assembly. The speed of iteration was measured because it could be so done at the same time as accuracy was being measured given a single tangent-stiffness calculation.  It was assumed that the order that each method was evaluated in parallel was unimportant, that evaluating each method successively within each solver iteration did not affect their performance individually and that speed of computation did not change with time. These assumptions allowed the test program to run the same problem with each of the methods at effectively the same time and generate an equal volume of data from each method. It was also assumed that the tangent-stiffness matrix calculation time for each of the methods did not vary based on the current configuration, as it changes slightly from iteration to iteration as the Newton solution iterate is updated.  This assumption allowed calculation time measurements to be averaged over all iterations within a single load step. The purpose of averaging calculation time measurements was to informally address the extraneous variables of parallel evaluation order and computer system load due to system processes not associated with the study. While not part of the study, the example programs available in the paper repository allow the reader to make a comparison of the methods for themselves on the basis of accuracy and convergence rate for two example nonlinear systems. In these examples, one will notice that the finite difference methods will lose quadratic convergence when the step size selected is too large such that accuracy relative to AD is decreased.

The goals of collecting accuracy and speed measurements were achieved by developing and implementing metrics within the simulation program used in the study. The metric used to measure the accuracy of a tangent-stiffness matrix was the Frobenius norm (analogous to an $l^2$-norm except defined on a matrix) of the element-wise difference between the tangent-stiffness matrix produced by the method being evaluated and the tangent-stiffness matrix produced by the AD based method (i.e., the \emph{exact} derivatives), given that both methods were set upon the same problem. The lower the value of this metric, the more accurate the method. The expression for the accuracy metric was
%
\begin{equation} 
  D = \sqrt{\sum_i \sum_j(K^{AD}_{ij} - K^{M}_{ij})^2}
  \label{eqn:accuracy}
\end{equation}
%
Where $D$ is distance, $K^{AD}$ is the tangent-stiffness matrix produced by the AD based method, $M$ is replaced with either CS, FD, or CD depending on the method being compared.  The metric used to compare the speed of the different methods was the time in seconds required to calculate the tangent-stiffness matrix. A final basis of comparison used in the study called \emph{computational efficiency} is based on specific calculation time per tangent-stiffness matrix element.

\subsubsection{Computers and other software} 
%
Discretization for the test problems run in the study was done using the software package \emph{Cubit} developed at Sandia National Laboratories \cite{ref-Cubit}. \emph{Cubit} generates a finite element mesh that is internally converted to a ``particle'' discretization internally when calling \emph{Peridigm}. The number of computational nodes in \emph{Peridigm} correspond to the number of finite elements in the discretization, not the number of finite element nodes.  An example journal script can be found on the corresponding author's website.

The test problems were run on the \emph{Stampede} HPC cluster computer housed at \emph{TACC} (Texas Advanced Computing Center) at The University of Texas at Austin. The single core test series was run using the {\tt normal} queue compute nodes, while the parallel test series was run using one to four {\tt large memory} nodes each having one terabyte of local RAM. This large amount of memory was required for reasons explained in Section~\ref{JGAM}.

\subsubsection{Data Reduction} 
The output produced by \emph{Perdigm} including the additional accuracy and speed measurements was redirected from the console to text files by the resource manager, \emph{SLURM} (Simple Linux User Resource Manager).  Directories and filenames were chosen so that data would be easy to sort by individual test run.  \emph{Python} scripts were then used to post-process and plot the data. These scripts averaged accuracy or speed data for a tests corresponding to a particular peridynamic node density over all iterations for that run, and then plotted this averaged data  as a function of peridynamic node density as it varied from test to test in the series.  Similarly, for the parallel tests, the data was averaged and plotted in much the same way, but as a function of number of compute cores used to solve the problem. Confidence intervals were not calculated on the measurements. The main reason is that network traffic from other users makes running tests on the HPC cluster a time varying process. These data reduction and plotting scripts can be found on the corresponding author's website.
